Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-05-31 main.tex
Merged in latest corrections from Allen.

% Pages are numbered in submission mode, and unnumbered in camera-ready
\ifcvprfinal\pagestyle{empty}\fi
\begin{document}

%\title{Accelerating $\ell_1$-Minimization Using Many-Core CPUs/GPUs \\ and Application to Face Recognition
%\title{Efficient Parallelization of Sparse Representation for Face Recognition}
%\thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}}
%\title{Many-Core Parallelization of $\ell_1$-Minimization for Face Recognition
\title{Parallelization of Fast $\ell_1$-Minimization for Face Recognition}

@@

\begin{abstract} 
Recently a family of promising face recognition algorithms that are based on sparse representation and $\ell_1$ minimization have been been developed.  These algorithms have not yet seen commercial application, largely due to their greater computational requirements compared to more traditional algorithms.  This paper studies techniques for leveraging the massive parallelism available in GPU and multi-core CPU hardware to accelerate augmented Lagrangian method (ALM) based $\ell_1$ minimization solvers.  The proposed implementations are significantly faster than than previously published implementations.  Despite the low arithmetic intensity of ALM solvers, the GPU can still outperform the CPU by a wide margin for large problem sizes, and when solving many small problems simultaneously.  The proposed CPU implementation remains faster for solving a single moderately sized problem.  
\end{abstract}

\section{Introduction} 
$\ell_1$ minimization ($\ell_1$-min) has received much attention in recent
years due to important applications in compressive sensing
\cite{BrucksteinA2007} and sparse representation \cite{WrightJ2010-PIEEE}.  
$\ell_1$-min refers to finding the minimum $\ell_1$-norm solution to an
underdetermined linear system $\bb=A\xx$:
\begin{equation}
\min \|\xx\|_1\quad \mbox{ subj. to }\quad \bb = A\xx.
\label{eq:l1min}
\end{equation}
It is now well established that, under certain conditions
\cite{CandesE2005-IT_1,DonohoD2004}, the minimum $\ell_1$-norm solution is also
the \emph{sparsest} solution to the system \eqref{eq:l1min}.

@@

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success also relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to iteratively align
a query image to each subject class individually. The face alignment algorithm solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain.  The sequence
$\tau_i$ converges to a transformation that aligns the test image $\tilde\bb$
with the training images from the $i$-th class $A_i$. In other words, the algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} to use the
$\ell_1$-norm as a robust error function, while simultaneously estimating
the coefficients of the linear illumination model $A\x$.

@@

In this discussion we will clearly distinguish between algorithm {\em parallelism},
which is a property of an algorithm, and hardware {\em concurrency}, which is a
property of a hardware architecture. Parallelism in an algorithm provides the
opportunity to perform computations concurrently on hardware.  Algorithms often
exhibit multiple levels of parallelism, and hardware often provides multiple
levels of concurrency.  In these terms, the primary contribution of this paper
will be to determine the optimal mappings between the parallelism available in
$\ell_1$-min and the concurrency available in multi-core CPU and GPU
architectures.

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of users or more, will likely require computation be
distributed over a cluster or network of computers, and this is not addressed here.
The source code of our implementations, which target Nvidia's CUDA GPU architecture,
as well as the x86-64 architecture, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

\subsection{Literature Review} 
We first briefly review the literature of $\ell_1$-min and existing
works on its parallelization.  Traditionally, $\ell_1$-min (or the so-called
basis pursuit (BP)) has been formulated as a linear program
\cite{ChenS2001-SIAM}. Several variations of the solution are also well known
in optimization, including a noisy approximation via quadratic programming
called the LASSO \cite{TibshiraniR1996} and truncated Newton interior-point
method (TNIPM) \cite{KimS2007}.

@@

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves
computationally expensive operations such as matrix-matrix multiplication and
the solution of least-squares problems with varying $A$
matrices.\footnote{During the face alignment algorithm presented later, The $A$
matrix changes between alignment iterations but not within the L1 solver}  In
Section \ref{sec:ALM}, we contend that ALM is a better choice for
implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.  IST
algorithms mainly involve elementary operations such as vector algebra and
matrix-vector multiplication. Therefore, when the dimension of the problem
becomes high, IST-type algorithms are particularly suitable for hardware
systems with a high degree of concurrency. In \cite{YangA2010-ICIP}, the
authors showed that ALM is able to significantly improve the solver speed,
while achieving estimation accuracy competitive with exact $\ell_1$-min
solutions such as BP and Homotopy methods, and other IST-type
algorithms.  Therefore, we choose ALM as the core algorithm for
implementation of the parallel face recognition pipeline.

@@

\section{Augmented Lagrange Multiplier Method}
\label{sec:ALM}

For the completeness of the paper, in this section, we briefly describe the ALM
algorithm for $\ell_1$-min \eqref{eq:l1min} \cite{YangA2010-ICIP}
 %and analyze its complexity
. 
Parallelization of ALM for face alignment and recognition will
be discussed in Section \ref{sec:alignment} and Section \ref{sec:recognition},
respectively.

Lagrange multiplier methods have been frequently used in convex programming to
eliminate equality constraints via adding a significant penalty term to the
cost function for infeasible points. ALM methods differ from other
penalty-based approaches by simultaneously estimating the optimal solution and
Lagrange multipliers in an iterative fashion.  For $\ell_1$-min, the augmented
Lagrangian is defined as: \begin{equation} L_\mu(\xx,\yy) = \|\xx\|_1 +
\yy^T(\bb - A\xx - \ee) + \frac{\mu}{2} \| \bb-A\xx-\ee \|_2^2, \end{equation}
where $\mu > 0$ is a constant that penalizes infeasibility and $\yy$ is a
vector of lagrange multipliers.

In Lagrange Multiplier Theory, if there exists a Lagrangian $y^*$ that
satisfies the second-order sufficiency conditions for optimality
\cite{BertsekasD2003}, then for a sufficiently large $\mu$, the optimal
$\ell_1$-min solution also minimizes
\begin{equation}
\xx^* = \arg \min L_\mu(\xx,\yy^*).
\label{eq:optimal-ALM}
\end{equation}

In practice, the optimal values for the triplet $(\xx^*, \yy^*, \mu)$ are all
unknown. Furthermore, it has been observed that solving
\eqref{eq:optimal-ALM} with a large initial value of $\mu$ tends to lead to
slower convergence speed \cite{WrightS2008,YangA2010-ICIP}. In
\cite{BertsekasD2003,YangJ2009}, an alternating procedure has been shown to
iteratively update $\xx$ and $\yy$:
\begin{equation}
\left \{
\begin{array}{lll}
\xx_{k+1} & = & \arg\min_{\xx} \, L_{\mu_k} (\xx,\yy_k)\\
\yy_{k+1} & = & \yy_k + \mu_k (\bb - A\xx_{k+1}) \\
\end{array}
\right . ,
\label{eq:alm}
\end{equation}
where $\mu_{k}\rightarrow \infty$ is a monotonically increasing positive sequence.
The iteration terminates when the estimates $(\xx, \yy)$ converge.

@@

{\small
\bibliographystyle{ieee}
\bibliography{faces}
}

>>>>

% Pages are numbered in submission mode, and unnumbered in camera-ready

\hyphenation{under-determined}
\ifcvprfinal\pagestyle{empty}\fi
\begin{document}

%\title{Accelerating $\ell_1$-Minimization Using Many-Core CPUs/GPUs \\ and Application to Face Recognition
%\title{Efficient Parallelization of Sparse Representation for Face Recognition}
%\thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}}
%\title{Many-Core Parallelization of $\ell_1$-Minimization for Face Recognition
\title{Parallelization of Fast $\ell_1$-Minimization for Face Recognition}
%\title{Parallelization of Fast $\ell_1$-Minimization Towards Real-Time Face Recognition}

@@

\begin{abstract} Recently a family of promising face recognition algorithms
based on sparse representation and $\ell_1$-minimization ($\ell_1$-min) have been
developed.  These algorithms have not yet seen commercial application, 
largely due to higher computational cost compared to other traditional
algorithms. This paper studies techniques for leveraging the 
massive parallelism available in GPU and CPU hardware to accelerate
$\ell_1$-min based on augmented Lagrangian method (ALM) solvers. We demonstrate
through extensive simulation and real-world experiment that the proposed
algorithms significantly speed up the sparsity recovery problem compared to the
state of the art.  Despite the low arithmetic intensity of ALM solvers, the GPU
can still outperform the CPU by a wide margin for large problem sizes, and when
solving many small problems simultaneously. The CPU implementation remains
faster for solving a single moderately sized problem. The source code of the
algorithms will be made available for peer evaluation.  \end{abstract}

\section{Introduction} 
$\ell_1$-minimization ($\ell_1$-min) has received much attention in recent
years due to important applications in compressive sensing
\cite{BrucksteinA2007} and sparse representation \cite{WrightJ2010-PIEEE}.  
The problem refers to finding the minimum $\ell_1$-norm solution to an
underdetermined linear system $\bb=A\xx$:
\begin{equation}
\min \|\xx\|_1\quad \mbox{ subj. to }\quad \bb = A\xx.
\label{eq:l1min}
\end{equation}
It is now well established that, under certain conditions
\cite{CandesE2005-IT_1,DonohoD2004}, the minimum $\ell_1$-norm solution is also
the \emph{sparsest} solution to the system \eqref{eq:l1min}.

@@

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success also relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to iteratively align
a query image to each subject class individually. The face alignment algorithm solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain.  The sequence
$\tau_i$ converges to a transformation that aligns the test image $\tilde\bb$
with the training images from the $i$-th class $A_i$. In other words, the algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} to use the
$\ell_1$-norm as a robust error function, while simultaneously estimating
the coefficients of the linear illumination model $A_i\x$.

@@

To clarify discussion, we clearly distinguish between {\em parallelism},
which is a property of an algorithm, and {\em concurrency}, which is a
property of a hardware architecture. Parallelism in an algorithm provides the
opportunity to perform computations concurrently on hardware.  Algorithms often
exhibit multiple levels of parallelism, and hardware often provides multiple
levels of concurrency.  In these terms, the primary contribution of this paper
will be to determine the optimal mappings between the parallelism available in
$\ell_1$-min and the concurrency available in multi-core CPU and GPU
architectures.

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of users or more, will likely require computation be
distributed over a cluster or network of computers, and this is not addressed here.
The source code of our implementations, which target Nvidia's CUDA GPU architecture
as well as the x86-64 architecture, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

\subsection{Literature Review} 
Traditionally, $\ell_1$-min (a.k.a.
basis pursuit (BP)) has been formulated as a linear program
\cite{ChenS2001-SIAM}. Several variations of the solution are also well known
in optimization, including a noisy approximation via quadratic programming
called the LASSO \cite{TibshiraniR1996} and truncated Newton interior-point
method (TNIPM) \cite{KimS2007}.

@@

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves
computationally expensive operations such as matrix-matrix multiplication and
linear least-squares problems with varying $A$
matrices.
%\footnote{During the face alignment algorithm presented later, The $A$ matrix changes between alignment iterations but not within the L1 solver}  
In Section \ref{sec:ALM}, we contend that ALM is a better choice for
implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.  IST
algorithms mainly involve elementary operations such as vector algebra and
matrix-vector multiplication. Therefore, when the dimension of the problem
becomes high, IST-type algorithms are particularly suitable for hardware
systems with a high degree of concurrency. In \cite{YangA2010-ICIP}, the
authors showed that ALM is able to significantly improve the solver speed,
while achieving estimation accuracy competitive with exact $\ell_1$-min
solutions such as BP and Homotopy methods, and other IST-type
algorithms. Therefore, we choose ALM as the core algorithm for
implementation of $\ell_1$-min in the parallel face recognition pipeline.

@@

\section{Augmented Lagrangian Method}
\label{sec:ALM}

In this section, we briefly describe the ALM algorithm for $\ell_1$-min \eqref{eq:l1min} \cite{YangA2010-ICIP}
and analyze its complexity. Lagrange multipliers have been frequently used in convex programming to
eliminate equality constraints via adding a penalty term to the
cost function for infeasible points. ALM methods differ from other
penalty-based approaches by simultaneously estimating the optimal solution and
Lagrange multipliers in an iterative fashion.  For $\ell_1$-min \eqref{eq:l1min}, the augmented
Lagrange function is defined as: 
\begin{equation} L_\mu(\xx,\yy) = \|\xx\|_1 +
\yy^T(\bb - A\xx) + \frac{\mu}{2} \| \bb-A\xx \|_2^2, 
\end{equation}
where $\mu > 0$ is a constant that penalizes infeasibility and $\yy$ is a
vector of Lagrange multipliers.

In Lagrange Multiplier Theory \cite{BertsekasD2003}, if there exists a Lagrangian $y^*$ that
satisfies the second-order sufficiency conditions for optimality, then for a sufficiently large $\mu$, the optimal
$\ell_1$-min solution also minimizes
\begin{equation}
\xx^* = \arg \min L_\mu(\xx,\yy^*).
\label{eq:optimal-ALM}
\end{equation}

In practice, the optimal values for the triplet $(\xx^*, \yy^*, \mu)$ are all
unknown. Furthermore, it has been observed that solving
\eqref{eq:optimal-ALM} with a large initial value of $\mu$ tends to lead to
slower convergence speed \cite{WrightS2008,YangA2010-ICIP}. In
\cite{BertsekasD2003,YangJ2009}, an alternating procedure has been shown to
iteratively update $\xx$ and $\yy$:
\begin{equation}
\left \{
\begin{array}{lll}
\xx_{k+1} & = & \arg\min_{\xx} \, L_{\mu_k} (\xx,\yy_k)\\
\yy_{k+1} & = & \yy_k + \mu_k (\bb - A\xx_{k+1}) \\
\end{array}
\right . ,
\label{eq:alm}
\end{equation}
where $\mu_{k}\rightarrow \infty$ is monotonically increasing.
The iteration terminates when the estimates $(\xx_k, \yy_k)$ converge.

@@

{\small
\bibliographystyle{ieee}
\bibliography{faces-short}
}
